Responses to Review Comments

### 2025-06-24 writeUps/paper.tex
edits

This paper expands on those latter results by presenting a new principal effect estimator, that boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.

After the initial submission of this manuscript, we became aware of \citet{richardson2023estimating}, which suggests a similar, though not identical, estimator to ours, using G-estimation of a structural nested mean model \citep[e.g.][]{wallace2017r}.
Our development of \textsc{geepers} builds on \citet{richardson2023estimating} by emphasizing connections to both parametric mixture modeling and principal scores analysis, and comparing \textsc{geepers} to those methods in a simulation study.
Moreover, \textsc{geepers}, which, in practice, is a combination of logistic regression and OLS, is more accessible to applied researchers while yielding very similar results.

@@

We will focus on estimation in completely randomized experiments, that is, we will assume
\begin{ass}[Randomization]\label{ass:rand}
\begin{equation*}%\label{eq:randomization}
  \yci,\yti,\sti, \bm{x}_i \independent Z_i
\end{equation*}
\end{ass}
where $\bm{x}_i$ is a vector of pre-treatment covariates, and $\independent$ denotes independence.

@@

\subsubsection{Parametric Mixture Modeling}
%The normal mixture modeling approach to estimating $\muc0$ and $\muc1$ \citep{imbens1997bayesian} assumes that, conditional on $\st$, $Y_C$ is normally distributed.
Say an analyst assumes that, within principal strata, $Y_C$ is drawn from a parametric model with parameter vector $\bm{\delta}^s$, i.e., $f_{Y_C}(y|\st=s)=g^s(y)$. For instance, \citep{imbens1997bayesian} assumes $Y_C$ is normally distributed within strata, so $g^s(y)=\phi\left(\frac{y-\mucs}{\sigma_C^s}\right)$, where $\phi(\cdot)$ is the standard normal density function, and $\sigma_C^0>0$ and $\sigma_C^1>0$ are standard deviations. 
Then the probability density of $Y_C$ may be written as
\begin{equation}\label{eq:mixtureModel}
 % f_{Y_C}(y)=Pr(\st=0)\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+Pr(\st=1)\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
 f_{Y_C}(y)=Pr(\st=0)g^0(y)+Pr(\st=1)g^1(y).
\end{equation}
The probabilities $Pr(\st=0)$ and $Pr(\st=1)$ may be estimated first using data from the treatment group, since Assumption \ref{ass:rand} implies that $Pr(\st=1|Z=1)=Pr(\st=1)$.
Given these estimates and \eqref{eq:mixtureModel}, parameters $\muc0$ and $\muc1$ may be estimated using maximum likelihood, method of moments, or Bayesian techniques.

@@

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi),
\end{equation}
%where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
the probability of a subject with covariates $\bxsi$ being in the $\sti=1$ principal stratum.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=0$.

@@

Principal scores can potentially improve inference in a finite mixture model, by re-writing \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y|\bm{x})=(1-\ppip{\bm{\alpha}})%\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)
g^0(y|\bm{x})+\ppip{\bm{\alpha}}%\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
g^1(y|\bm{x})
\end{equation}
where $f_{Y_C}(y|\bm{x})$ and $g^s(y|\bm{x})$, $s=0,1$, are conditional densities which must be specified. Most commonly, researchers assume a linear normal model, i.e., $g^s(y|\bm{x})=\phi\left(\frac{y-\mucs-\bm{\gamma}'\bm{x}}{\sigma_C^s}\right)$, where the columns of $\bm{x}$ are centered so that $\mathbb{E}[\bm{x}]=\bm{0}$. 
%Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

Alternatively, an analyst may assume ``principal ignorability'' \citep{jo,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies. % \citep[e.g.][]{rosenbaum2010design}.
% Both require a rich set of covariates and background knowledge of the determinants of $S$;
In particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

Under principal ignorability, $\muc0$ and $\muc1$ can be estimated via weighting:
\begin{equation}\label{eq:psw}
  \hat{\muc0}_w=\frac{\sum_{i: Z_i=0} Y_i(1-\pp)}{\sum_{i:Z_i=0}1-\pp}\text{ and } \hat{\muc1}_w=\frac{\sum_{i: Z_i=0} Y_i(\pp)}{\sum_{i:Z_i=0}\pp}
\end{equation}

@@

Multiple M-estimators can be combined by ``stacking'' their estimating equations.
In particular, say an M-estimator with estimating equations $\bm{\psi}^2(\bm{\mathcal{O}},\bm{\hat\theta^2})=\bm{0}$ takes, as input, the results from a previous M-estimator, with estimating equations $\bm{\psi}^1(\bm{\mathcal{O}},\bm{\hat\theta^1})=\bm{0}$.
The resulting estimator is the solution to the combined estimating equations
\begin{equation}\label{eq:stacked}
  \psi(\bm{\mathcal{O}},[\bm{\hat\theta^1},\bm{\hat{\theta}^2}])=[\psi^1(\bm{\mathcal{O}},\bm{\hat\theta^1}),\psi^2(\bm{\mathcal{O}},\bm{\hat\theta^2})]=\bm{0},
\end{equation}
%where the superscript $s$ stands for ``stacked.''
A variance-covariance matrix estimated using the sandwich formula based on $\psi^s$ accounts for uncertainty from both steps of the estimation procedure.

@@

The methods all used the same ``observed'' dataset, consisting of outcomes $Y$, treatment assignments $Z$, two covariates $\bxs=\bxy=\bx=[x_1,x_2]$, and $S$ observed only when $Z=1$.
The third covariate used in the data generating model, $x_3$, was ``unobserved''--since $x_3$ was correlated with both $\st$ and $Y$, Principal Ignorability (Assumption \ref{ass:PI}) was violated.

>>>>

This paper expands on those latter results by presenting a new principal effect estimator for the case of one-way noncompliance, in which principal stratum membership is observed in one treatment group but not the other (future work will extend our approach to general scenarios). Our estimator boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.

After the initial submission of this manuscript, we became aware of \citet{richardson2023estimating}, which suggests a similar, though not identical, estimator to ours, using G-estimation of a structural nested mean model \citep[e.g.][]{wallace2017r}.
Our development of \textsc{geepers} builds on \citet{richardson2023estimating} by emphasizing connections to both parametric mixture modeling and principal score analysis, and comparing \textsc{geepers} to those methods in a simulation study.
Moreover, \textsc{geepers}, which, in practice, is a combination of logistic regression and OLS, is more accessible to applied researchers while yielding very similar results.

@@

We will focus on estimation in completely randomized experiments; that is, we will assume
\begin{ass}[Randomization]\label{ass:rand}
\begin{equation*}%\label{eq:randomization}
  \yci,\yti,\sti, \bm{x}_i \independent Z_i
\end{equation*}
\end{ass}
where $\bm{x}_i$ is a vector of pre-treatment covariates, and $\independent$ denotes independence.

@@

\subsubsection{Parametric Mixture Modeling}
%The normal mixture modeling approach to estimating $\muc0$ and $\muc1$ \citep{imbens1997bayesian} assumes that, conditional on $\st$, $Y_C$ is normally distributed.
Say an analyst assumes that, within principal strata, $Y_C$ is drawn from a parametric model, i.e., $f_{Y_C}(y|\st=s)=g^s(y)$. For instance, \citep{imbens1997bayesian} assumes $Y_C$ is normally distributed within strata, so $g^s(y)=\phi\left(\frac{y-\mucs}{\sigma_C^s}\right)$, where $\phi(\cdot)$ is the standard normal density function, and $\sigma_C^0>0$ and $\sigma_C^1>0$ are standard deviations. 
Then the probability density of $Y_C$ may be written as
\begin{equation}\label{eq:mixtureModel}
 % f_{Y_C}(y)=Pr(\st=0)\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+Pr(\st=1)\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
 f_{Y_C}(y)=Pr(\st=0)g^0(y)+Pr(\st=1)g^1(y).
\end{equation}
The probabilities $Pr(\st=0)$ and $Pr(\st=1)$ may be estimated first using data from the treatment group, since Assumption \ref{ass:rand} implies that $Pr(\st=1|Z=1)=Pr(\st=1)$.
Given these estimates and \eqref{eq:mixtureModel}, parameters $\muc0$ and $\muc1$ may be estimated using maximum likelihood, method of moments, or Bayesian techniques.

@@

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi),
\end{equation}
%where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
the probability of a subject with covariates $\bxsi$ being in the $\sti=1$ principal stratum.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ as $\bm{\hat{\alpha}}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\hppi=\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=0$.

@@

Principal scores can potentially improve inference in a finite mixture model, by re-writing \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y|\bm{x})=(1-\ppip{\bm{\alpha}})%\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)
g^0(y|\bm{x})+\ppip{\bm{\alpha}}%\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
g^1(y|\bm{x})
\end{equation}
where $g^s(y|\bm{x})$, $s=0,1$, are conditional densities which must be specified. Most commonly, researchers assume a linear normal model, e.g., $g^s(y|\bm{x})=\phi\left(\frac{y-\mucs-\bm{\gamma}'\bm{x}}{\sigma_C^s}\right)$, where the columns of $\bm{x}$ are centered so that $\mathbb{E}[\bm{x}]=\bm{0}$. 
%Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

Alternatively, an analyst may assume ``principal ignorability'' \citep{jo,dingLu}:
%\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
%\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies. % \citep[e.g.][]{rosenbaum2010design}.
% Both require a rich set of covariates and background knowledge of the determinants of $S$;
In particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

Under principal ignorability, $\muc0$ and $\muc1$ can be estimated via weighting:
\begin{equation}\label{eq:psw}
  \hat{\muc0}_w=\frac{\sum_{i: Z_i=0} Y_i(1-\ppi)}{\sum_{i:Z_i=0}1-\ppi}\text{ and } \hat{\muc1}_w=\frac{\sum_{i: Z_i=0} Y_i(\ppi)}{\sum_{i:Z_i=0}\ppi}
\end{equation}

@@

Multiple M-estimators can be combined by ``stacking'' their estimating equations.
In particular, say an M-estimator with estimating equations $\bm{\psi}^2(\bm{\mathcal{O}},\bm{\hat\theta^2})=\bm{0}$ takes, as input, the results from a previous M-estimator, with estimating equations $\bm{\psi}^1(\bm{\mathcal{O}},\bm{\hat\theta^1})=\bm{0}$.
The resulting estimator is the solution to the combined estimating equations
\begin{equation}\label{eq:stacked}
  \psi(\bm{\mathcal{O}},[\bm{\hat\theta^1},\bm{\hat{\theta}^2}])=[\psi^1(\bm{\mathcal{O}},\bm{\hat\theta^1}),\psi^2(\bm{\mathcal{O}},\bm{\hat\theta^2})]=\bm{0},
\end{equation}
%where the superscript $s$ stands for ``stacked.''
A variance-covariance matrix estimated using the sandwich formula based on $\psi$ accounts for uncertainty from both steps of the estimation procedure.

@@

The methods all used the same ``observed'' dataset, consisting of outcomes $Y$, treatment assignments $Z$, two covariates $\bxs=\bxy=\bx=[x_1,x_2]$, and $S$ observed only when $Z=1$.
The third covariate used in the data generating model, $x_3$, was ``unobserved''--since $x_3$ was correlated with both $\st$ and $Y$, Principal Ignorability \eqref{eq:pi} was violated.

### 2025-06-24 writeUps/responses.tex
edits

\item \begin{itquote}      APA Citations and References: The paper does not adhere to the APA 7th style. Authors are advised to consult the APA Formatting and Style Guide for proper citation and reference formatting throughout the manuscript. Consistency in style is crucial for scholarly publications.
\end{itquote}
Thank you. We have fixed the formatting.
\end{enumerate}

@@

\item \begin{itquote}   Please provide more background on M-estimators especially to the extent it evolves to become relevant to the GEEPERs. 
\end{itquote}
Thanks for this suggestion---in response, we added Section \ref{sec:mest}
\item \begin{itquote}      I am wondering if doubly robust PSW estimator would reduce the remaining bias difference between GEEPERs and PSW, and Mixture and PSW.
\end{itquote}
This is an interesting suggestion---you're right that the \psw estimator did not include an outcome model, unlike \geepers and \nmm. However, we doubt that an outcome model would substantially reduce the bias of \psw in the simulation. The advantage of doubly-robust estimation is that an outcome model can correct for model misspecification in the propensity (or principal) score model. However, the principal score model in our simulation study was well-specified---the bias of the \psw estimator was due to an unobserved confounder, $x_3$. Unfortunately, like \psw, doubly-robust estimators rely on ignorability, i.e., that all relevant confounders have been measured and included. 
\item \begin{itquote}      GEEPERs seems great, but also, surprised by how efficient Mixture can be. Are there any mechanisms that link M-estimators to Mixture modeling? Just wondering.
\end{itquote}
We were also surprised by how well the \nmm performed in the simulation, especially given prior literature detailing its vulnerability \citep[][; cited in the paper]{griffin2008application,feller2016principal}. Its similarity to \geepers in some scenarios is less surprising, since they are both essentially mixture model estimators.

>>>>

\item \begin{itquote}      APA Citations and References: The paper does not adhere to the APA 7th style. Authors are advised to consult the APA Formatting and Style Guide for proper citation and reference formatting throughout the manuscript. Consistency in style is crucial for scholarly publications.
\end{itquote}
Thank you. We have fixed the formatting.
\End{enumerate}

@@

\item \begin{itquote}   Please provide more background on M-estimators especially to the extent it evolves to become relevant to the GEEPERs. 
\end{itquote}
Thanks for this suggestion---in response, we added Section \ref{sec:mest}
\item \begin{itquote}      I am wondering if doubly robust PSW estimator would reduce the remaining bias difference between GEEPERs and PSW, and Mixture and PSW.
\end{itquote}
This is an interesting suggestion---you're right that the \psw estimator did not include an outcome model, unlike \geepers and \pmm. However, we doubt that an outcome model would substantially reduce the bias of \psw in the simulation. The advantage of doubly-robust estimation is that an outcome model can correct for model misspecification in the propensity (or principal) score model. However, the principal score model in our simulation study was well-specified---the bias of the \psw estimator was due to an unobserved confounder, $x_3$. Unfortunately, like \psw, doubly-robust estimators rely on ignorability, i.e., that all relevant confounders have been measured and included. 
\item \begin{itquote}      GEEPERs seems great, but also, surprised by how efficient Mixture can be. Are there any mechanisms that link M-estimators to Mixture modeling? Just wondering.
\end{itquote}
We were also surprised by how well the \pmm performed in the simulation, especially given prior literature detailing its vulnerability \citep[][; cited in the paper]{griffin2008application,feller2016principal}. Its similarity to \geepers in some scenarios is less surprising, since they are both essentially mixture model estimators.
